---
title: "Nationality-Based Discrimination in Mexico's Refugee Determination Process"
author: "Juan C. Venancio"
description: Gov 50 Immigration Project
output:
  html_document:
    df_print: paged
distill::distill_article:
  self_contained: no
---



## Introduction

The 1951 United Nations Refugees Convention establishes the fundamental framework of international asylum law for member states of the United Nations. Since 1951, member states have met in other conventions to re-contextualize and re-interpret this legal framework in new correspondence to their national interests and needs (OAU Convention 1969 & 1984 Cartagena Declaration), but the legal principles establishing the definition of a refugee remain identical since its conception. According to the 1951 Convention, a refugee is any person that based on a well-founded fear of persecution on the basis of their race, religion, nationality, membership to a particular social group, or political opinion, merits the international protection of a sovereign entity outside of their country of origin. As it stands, in 2022, approximately 108.4 million people were forcibly displaced, and of those forcibly displaced, 35.3 million were categorized as refugees. According to the UNHCR, 76% of the world’s refugees are hosted in low and middle-income countries across the world. 

Since its inception, Mexico has not historically been a destination country for migrants and refugees. At most, the country became a transit country for Central and South American migrants that headed to the United States during periods of increased generalized drug violence, civil wars, and economic destitution between the late 1980s and early 2000s. Over the last decade alone, the number of refugee applications submitted to and processed by the Mexican government has exponentially increased, and this data is indicative of Mexico's evolving role as a destination country for many immigrants. However, given the exponential increase in applications and Mexico's limited institutional capacity as a middle-income country, questions have emerged whether COMAR, Mexico's refugee processing institution, selective prioritizes applications of certain nationalities and therefore discriminates against others. I would like to explore this phenomenon, and compile/analyze data that uncover potential discriminatory institutional practices on the basis of Hispanic and non-Hispanic categorizations, as well as suggest potential policy recommendations. I am planning to use data from the Mexican Commission for Refugee Assistance. 




## Project Abstract & Methodology

Research Question: To What Extent Does Hispanic Nationality Impact Refugee Status Determination
in Mexico?
  
Hypothesis: I hypothesize, that migrants who indicate a Hispanic nationality tend to have higher possibilities to successfully obtain refugee status in Mexico. As a result, I expect that migrants from Haiti, the non-Spanish Caribbean, and other parts of the world tend to have lower possibilities at refugee resettlement. Additionally, though the Mexican Government has made advances to expand their refugee granting criteria, because the 1951 UN Convention of Refugees provides a large number of successful reugee claims to migrants with "well-founded fear of persecution", then I also predict that Mexican institutions accept Spanish-speaking Latin-Americans at a significant larger rate. For this project, I will analyze the Mexican Government's Commission on Refugee Assistance open data set that I will use to assess the several reasons that migrants tend to leave their countries of origin. If Venezuelans, Hondurans, and other spanish speaking Latin Americans have a higher rate of refugee success, then that will make my hypothesis correct. If other types of nationalities have a higher rate of success, then I will be proven wrong. In any case, I am also interested in observing the variability of application success and the type of application submitted. 

Methodology: My unit of measurement is whether an immigrant was granted refugee status or not. If the immigrant was provided with a refugee status, implying that the migrant was considered to have had a well-founded fear of persecution, then I will indicate that with 1 for successful application. If a migrant was rejected, then I will indicate that with 0. The independent variable will be the migrant's nationality. The explanatory variable of interest is whether the nationality impacts the likelihood that a migrant gets refugee status in Mexico. 


```{r}
#set up all libraries needed for this project
library(tidyverse, ggplot2)

#import data needed
submitted_applications <- read.csv("//Users//jvenancio//Downloads//COMAR_Thesis_Data_Solicitudes.csv")
application_results <- read.csv("//Users//jvenancio//Downloads//COMAR_Thesis_Data_Resoluciones.csv")

```


```{r}
#count how many applications were submitted per nationality + create data table
apps_sub_per_nationality <- submitted_applications |> 
  count(NACIONALIDAD) |> 
  rename(applications = n, Nationality = NACIONALIDAD)

```

```{r}

#count how many applications were submitted per year + create data table
apps_sub_per_year <- submitted_applications |> 
  count(AÑO) |> 
  rename(applications = n, year = AÑO)

# Assuming 'submitted_applications' is your data frame




```






```{r}
#count how many applications were processed per year + create data table
apps_proc_per_year <- application_results |>
  count(AÑO_RESOLUCION) |>
  rename(applications = n, year = AÑO_RESOLUCION)

```

```{r}
#count the end result of the applications that were processed in total from 2013-23
apps_proc_result <- application_results |>
  count(SENTIDO_RESOLUCIÓN) 

# Print the data table
print(apps_sub_per_year)


```
## Results
```{r}
#count the end result of the applications that were processed (positive) in total from 2013-23 per nationality
apps_proc_result_pos <- application_results |>
  filter(SENTIDO_RESOLUCIÓN == "Resolución Positiva")

table(apps_proc_result_pos$NACIONALIDAD)

```
```{r}

apps_proc_result_neg <- application_results |>
  filter(SENTIDO_RESOLUCIÓN == "Resolución Negativa")

table(apps_proc_result_neg$NACIONALIDAD)






```


## Paragraph Summary: 
These two data pieces are essential to answer my research question. As a reminder, I am investigating whether the nationality of an applicant impacts the likelihood that they receive refugee status in Mexico. My hypothesis was that if an applicant is from a Spanish-speaking country, then they are more likely to acquire refugee status. If an applicant is from a non-Spanish-speaking country, then they are more likely to get rejected. 

Based on these data visuals, my hypothesis is correct. First, the nationality that is the most impacted is Haiti. According to my data set, while 2190 Haitians managed to acquire refugee status in Mexico, 10486 Haitian applicants have been rejected over the last decade. Haiti is a country in the Caribbean, whose official language is Haitian Creole. 

On the contrary, migrants from Venezuela have been overwhelmingly granted refugee status over the last decade. Only 882 Venezuelans were denied refugee status, and 16,366 Venezuelans were granted refugee status. Venezuela is a Spanish-speaking country in South-America. 

The reason why I specifically analyzed both cases is because arguably the biggest differences in their country's context are ethnic and cultural backgrounds. Both countries are experiencing the largest political instability contexts in contemporary Latin-American history, and are often impacted by economic and environmental destitution in the same. I would have expected that if nationality and the corresponding racial implications were not contributing determinants to refugee acquisition, then applicants from these two countries would have been admitted at statistically similar rates. However, despite their similar political and economic contexts, there is a drastically different outcome in refugee status determination in Mexico. 









```{r}
library(ggplot2)

# data frames: apps_sub_per_year, apps_proc_per_year, apps_proc_result

# Creating the line graph for submitted applications per year
ggplot(data = apps_sub_per_year %>% filter(year != 2023),
       aes(x = year, y = applications, group = 1)) +
  geom_line(color = "blue") +
  geom_point(color = "blue", size = 2) +
  geom_smooth(linetype = 2, se = FALSE) + 
  labs(title = "Submitted Refugee Applications in Mexico 2013-2023",
       x = "Year",
       y = "Number of Applications") +
  theme_minimal()

# Creating the line graph for processed applications per year
ggplot(data = apps_proc_per_year, aes(x = year, y = applications, group = 1)) +
  geom_line(color = "red") +
  geom_point(color = "red", size = 3) +
  labs(title = "Processed Refugee Applications in Mexico 2013-2023",
       x = "Year",
       y = "Number of Applications") +
  theme_minimal()

# Creating a bar plot for the result of processed applications
ggplot(data = apps_proc_result, aes(x = SENTIDO_RESOLUCIÓN, y = n)) +
  geom_bar(stat = "identity", fill = "black") +
  labs(title = "Result of Processed Asylum Applications in Mexico 2013-2023",
       x = "Result",
       y = "Number of Applications") +
  theme_minimal()




```






```{r}

library(ggplot2)


merged_data <- merge(apps_sub_per_year, apps_proc_per_year, by = "year", all = TRUE)


ggplot(data = merged_data, aes(x = as.factor(year))) +
  geom_bar(aes(y = applications.x, fill = "Submitted"), stat = "identity", position = "dodge", width = 0.4) +
  geom_bar(aes(y = applications.y, fill = "Processed"), stat = "identity", position = "dodge", width = 0.4) +
  labs(title = "Refugee Applications in Mexico (2013-2023)",
       x = "Year",
       y = "Number of Applications") +
  scale_fill_manual(name = "Application Type", values = c("Submitted" = "blue", "Processed" = "red")) +
  theme_minimal()



```



##

Paragraph: Since 2013, the number of refugee applications in Mexico have exponentially increased. Simultaneously, this bar graph indicates how Mexico has also processed, though not completely, refugee applications over the years. This graph indicates that Mexico has become a destination country, and every year there is an increasing number of migrnats that attempt to permanently resettle in the country. 






```{r}
library(dplyr)

spanish_speaking_countries <- c("Argentina","Chile","Colombia","Costa Rica","Cuba","Ecuador","El Salvador","EL SALVADOR","España","Guatemala","GUATEMALA","Guinea Ecuatorial","Honduras","Nicaragua","NICARAGUA","Panamá","Perú","República Dominicana","Uruguay","Venezuela","VENEZUELA")

data <- application_results |>
  mutate(spanish_speaking = if_else(NACIONALIDAD %in% spanish_speaking_countries, 1, 0),
         approved = if_else(SENTIDO_RESOLUCIÓN == "Resolución Positiva", 1, 0))

fit1 <- lm(approved ~ spanish_speaking, data = data)

summary(fit1)

```


```{r}

# Install and load necessary packages if not already installed
if (!requireNamespace("ggplot2", quietly = TRUE)) {
  install.packages("ggplot2")
}
if (!requireNamespace("dplyr", quietly = TRUE)) {
  install.packages("dplyr")
}

# Load libraries
library(ggplot2)
library(dplyr)

# Your linear regression model
model <- lm(formula = approved ~ spanish_speaking, data = data)

# Create a dataframe for predictions
predictions <- data.frame(data, predicted = predict(model))

# Plotting the data and the regression line
ggplot(predictions, aes(x = spanish_speaking, y = approved)) +
  geom_point() +  # Scatter plot of the data points
  geom_line(aes(x = spanish_speaking, y = predicted), color = "blue", size = 1) +  # Regression line
  labs(title = "Linear Regression: Correlation between Hispanic Nationality and Outcome",
       x = "Hispanic Nationality",
       y = "Positive Resolution Percentage") +
  theme_minimal()












```


```{r}
# Assuming 'application_results' is your data frame
library(dplyr)
library(knitr)

# Filter for positive resolutions
apps_proc_result_pos <- application_results %>%
  filter(SENTIDO_RESOLUCIÓN == "Resolución Positiva")

# Count applications per nationality for positive resolutions
table_result <- table(apps_proc_result_pos$NACIONALIDAD)

# Create a nice table
kable(data.frame(Nationality = names(table_result), Applications = as.vector(table_result)),
      col.names = c("Nationality", "Applications"),
      caption = "Number of Applications with Positive Resolutions Per Nationality")







```


```{R}
# Assuming 'application_results' is your data frame
library(dplyr)

# Count applications per positive resolution type
apps_proc_result_pos <- application_results %>%
  filter(SENTIDO_RESOLUCIÓN == "Resolución Positiva") %>%
  count(NACIONALIDAD)

# Count applications per negative resolution type
apps_proc_result_neg <- application_results %>%
  filter(SENTIDO_RESOLUCIÓN == "Resolución Negativa") %>%
  count(NACIONALIDAD)

# Merge the two tables
merged_table <- merge(apps_proc_result_pos, apps_proc_result_neg, by = "NACIONALIDAD", all = TRUE)

# Rename columns for clarity
colnames(merged_table) <- c("Nationality", "Positive_Resolutions", "Negative_Resolutions")

# Print the merged table
print(merged_table)



```

```{r}

# List of countries to filter
selected_countries <- c("Honduras", "El Salvador", "Guatemala", "Haití", "Brasil", "Federación de Rusia", "Ghana", "India", "Nicaragua", "Venezuela", "Nígeria")

# Filter the merged table for the selected countries
filtered_merged_table <- merged_table %>%
  filter(Nationality %in% selected_countries)

# Print or further process the filtered merged table
print(filtered_merged_table)





```

```{r}
# Assuming 'application_results' is your data frame
library(dplyr)

# Count applications per positive resolution type
apps_proc_result_pos <- application_results %>%
  filter(SENTIDO_RESOLUCIÓN == "Resolución Positiva") %>%
  count(NACIONALIDAD)

# Count applications per negative resolution type
apps_proc_result_neg <- application_results %>%
  filter(SENTIDO_RESOLUCIÓN == "Resolución Negativa") %>%
  count(NACIONALIDAD)

# Merge the two tables
merged_table <- merge(apps_proc_result_pos, apps_proc_result_neg, by = "NACIONALIDAD", all = TRUE)

# Rename columns for clarity
colnames(merged_table) <- c("Nationality", "Positive_Resolutions", "Negative_Resolutions")

# Add a percentage column
merged_table$Percentage_Positive <- paste0(round((merged_table$Positive_Resolutions / 
                                                   (merged_table$Positive_Resolutions + merged_table$Negative_Resolutions)) * 100, 1), "%")

# Filter out the total row
filtered_merged_table <- merged_table %>%
  filter(!is.na(Nationality))

# List of countries to filter
selected_countries <- c("Honduras", "El Salvador", "Guatemala", "Haití", "Brasil", "Federación de Rusia", "Ghana", "India", "Nicaragua", "Venezuela", "Nígeria")

# Filter the merged table for the selected countries
filtered_merged_table <- filtered_merged_table %>%
  filter(Nationality %in% selected_countries)

# Print or further process the filtered merged table
print(filtered_merged_table)



kable(filtered_merged_table, format = "html",
      col.names = c("Nationality", "Positive Resolutions", "Negative Resolutions", "Percentage of Positive Resolutions"),
      caption = "Refugee Application Outcomes by Top Hispanic and Non-Hispanic Nationalities")










```

```{r}
# Assuming 'application_results' is your data frame
library(dplyr)
library(knitr)

# Count applications per positive resolution type
apps_proc_result_pos <- application_results %>%
  filter(SENTIDO_RESOLUCIÓN == "Resolución Positiva") %>%
  count(NACIONALIDAD)

# Count applications per negative resolution type
apps_proc_result_neg <- application_results %>%
  filter(SENTIDO_RESOLUCIÓN == "Resolución Negativa") %>%
  count(NACIONALIDAD)

# Merge the two tables
merged_table <- merge(apps_proc_result_pos, apps_proc_result_neg, by = "NACIONALIDAD", all = TRUE)

# Rename columns for clarity
colnames(merged_table) <- c("Nationality", "Positive_Resolutions", "Negative_Resolutions")

# Add a percentage column (rounded to the nearest whole number)
merged_table$Percentage_Positive <- paste0(round((merged_table$Positive_Resolutions / 
                                                   (merged_table$Positive_Resolutions + merged_table$Negative_Resolutions)) * 100), "%")

# Filter out the total row
filtered_merged_table <- merged_table %>%
  filter(!is.na(Nationality))

# List of countries to filter
selected_countries <- c("Honduras", "El Salvador", "Guatemala", "Haití", "Brasil", "Federación de Rusia", "Ghana", "India", "Nicaragua", "Venezuela", "Nígeria")

# Filter the merged table for the selected countries
filtered_merged_table <- filtered_merged_table %>%
  filter(Nationality %in% selected_countries)
filtered_merged_table <- arrange(filtered_merged_table, desc(Percentage_Positive))

# Create a nice-looking table
kable(filtered_merged_table, format = "html",
      col.names = c("Nationality", "Positive Resolutions", "Negative Resolutions", "Percentage of Positive Resolutions"),
      caption = "Refugee Application Outcomes for Top Hispanic and Non-Hispanic Nationalities (2013-2023)")




```

```{r}
# Assuming 'filtered_merged_table' is your data frame
library(dplyr)
library(ggplot2)

# Filter the data for Venezuela and Haiti
venezuela_data <- filtered_merged_table %>%
  filter(Nationality == "Venezuela")

haiti_data <- filtered_merged_table %>%
  filter(Nationality == "Haití")

# Check if the filtered data frames are not empty
if (nrow(venezuela_data) > 0 && nrow(haiti_data) > 0) {
  # Combine the data for Venezuela and Haiti
  bar_data <- rbind(venezuela_data, haiti_data)

  # Create a side-by-side bar plot
  ggplot(bar_data, aes(x = Nationality, y = Positive_Resolutions, fill = Nationality)) +
    geom_bar(stat = "identity", position = "dodge") +
    labs(title = "Positive Resolutions of Venezuelan and Haitian Applicants",
         x = "Nationality",
         y = "Count") +
    scale_fill_manual(values = c("Venezuela" = "lightblue", "Haití" = "lightgreen")) +
    theme_minimal()
} else {
  cat("No data found for Venezuela or Haiti.\n")
}



```
